The error in Bayesian phylogenetic reconstruction when speciation is not instantaneous

### 2019-05-07 content.tex
No appendix, use 'Supplement'
Improved title

\richel{Start new}
The PBD model has five parameters, depicted in table \ref{table:parameters}. 
The per species speciation completion rates $\lambda$ 
we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
This means that there is a $0.15\thinspace dt$ 
($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
probability of speciation completion occurring in an infinitesimal time $dt$. 
We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
We use tree sizes of $n$ of 50, 100 and 200 good taxa.
For each combination of $\lambda$, $\mu$ and $n$, 
we use a speciation initiation rate $b = b_i = b_g$ 
so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
given a $b$, $\lambda$ and $\mu$,
equals the desired number of species $n$.
$b$ is calculated using the PBD R package (\cite{pbd})
for each parameter combination and shown in table \ref{table:pbd_parameters}.
We use $\lambda = 10^9 \approx \infty$ 
as our control for which the PBD model reduces to the BD model.

@@

We create two data sets: a general one, to explore parameter space, 
and one to investigate the effect of sampling incipient species (see below).
For the general data set, all the trees with the correct number of good species are kept.
There is an additional selection criterion, for the data set to investigate sampling:
to generate that data set, only incipient species trees are kept, 
on which the two sampling methods (see below) result in different species trees.
As sampling will never have an effect for $\lambda = \infty$, 
this parameter value is absent in that data set.
\richel{End new}

@@

\richel{Start new}
As described above, per alignment, we do four different Bayian analyses,
as we use two different site models and two different clock models.
We know the generative site model (which is JC69) and clock 
model (a strict clock), but want to explore how often the correct model
is indeed preferred. Per alignment, we measure the estimated marginal 
likelihood (which is the probability of the data given the model)
for each of the four models and measure their relative propertions.
To estimate the marginal likelihood, we use the novel 
Nested Sampling approach (\cite{maturana_et_al_2018}),
which we configure to have an effective sample size (in estimating
the marginal likelihood) of at least 200, by using a single-particle
sub-chain with a length equal to the MCMC interval.
\richel{
  There are no rules described in the setup of the NS
  (for example, what should the relation between MCMC chain length
  and NS sub-chain length be?),
  but one can experiment with a sub-chain length and number of particles.
  I will need to do some pilot studies to verify if this is
  the correct way to do and describe this.  
}
\richel{End new}

@@

\richel{Start new}
In this article, we showcase the effect of sampling
and the certainty when selecting a (site and clock) model. 
We show the effect sampling has on the inference error,
for the nLTT distributions generated from assuming the (correct) Jukes-Cantor 
site model and a strict clock model. We do so per parameter combination,
displaying the distribution as a violin plot.
In the supplementary information, we display these error distributions 
of the general data set all combined, or separated per site or clock model.
The certainty in the model selection is again shown per parameter setting,
as a stacked bar. 
\richel{
  I am unsure how to depict that stacked bar and its uncertainty.
  Sure, the stacked bars (without uncertainty) will work,
  but information about uncertainty is lost. 
  See figure 'Showing a stacked bar with uncertainty'
  for three equivalent ways to display a stacked bar with uncertainty.
}

@@

\richel{End new}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $b_g$                 & Speciation initiation rate of a good species & derived, see \ref{table:pbd_parameters} \\
    $b_i$                 & Speciation initiation rate of an incipient species & derived, see \ref{table:pbd_parameters} \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^9$ \\
    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2 \\
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_s$                 & Sampling method & S, L, R \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line is 
    the biological parameter set. 
    \richel{Start new}
    The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
    and so on.  
    The RNG seeds for the data set investigating the effect 
    of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
    \richel{End new}
    The sampling methods are abbreviated as such: 'R' denotes random
    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
    data set. For the data set exploring the effect of sampling, we use 'shortest'
    and 'longest' for each value of $R_i$ (which are random seeds in which sampling has an effect).
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

The PBD model has five parameters, depicted in table \ref{table:parameters}. 
The per species speciation completion rates $\lambda$ 
we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
This means that there is a $0.15\thinspace dt$ 
($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
probability of speciation completion occurring in an infinitesimal time $dt$. 
We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
We use tree sizes of $n$ of 50, 100 and 200 good taxa.
For each combination of $\lambda$, $\mu$ and $n$, 
we use a speciation initiation rate $b = b_i = b_g$ 
so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
given a $b$, $\lambda$ and $\mu$,
equals the desired number of species $n$.
$b$ is calculated using the PBD R package (\cite{pbd})
for each parameter combination and shown in table \ref{table:pbd_parameters}.
We use $\lambda = 10^9 \approx \infty$ 
as our control for which the PBD model reduces to the BD model.

@@

We create two data sets: a general one, to explore parameter space, 
and one to investigate the effect of sampling incipient species (see below).
For the general data set, all the trees with the correct number of good species are kept.
There is an additional selection criterion, for the data set to investigate sampling:
to generate that data set, only incipient species trees are kept, 
on which the two sampling methods (see below) result in different species trees.
As sampling will never have an effect for $\lambda = \infty$, 
this parameter value is absent in that data set.

@@

As described above, per alignment, we do four different Bayian analyses,
as we use two different site models and two different clock models.
We know the generative site model (which is JC69) and clock 
model (a strict clock), but want to explore how often the correct model
is indeed preferred. Per alignment, we measure the estimated marginal 
likelihood (which is the probability of the data given the model)
for each of the four models and measure their relative propertions.
To estimate the marginal likelihood, we use the novel 
Nested Sampling approach (\cite{maturana_et_al_2018}),
which we configure to have an effective sample size (in estimating
the marginal likelihood) of at least 200, by using a single-particle
sub-chain with a length equal to the MCMC interval.
\richel{
  There are no rules described in the setup of the NS
  (for example, what should the relation between MCMC chain length
  and NS sub-chain length be?),
  but one can experiment with a sub-chain length and number of particles.
  I will need to do some pilot studies to verify if this is
  the correct way to do and describe this.  
}

@@

In this article, we showcase the effect of sampling
and the certainty when selecting a (site and clock) model. 
We show the effect sampling has on the inference error,
for the nLTT distributions generated from assuming the (correct) Jukes-Cantor 
site model and a strict clock model. We do so per parameter combination,
displaying the distribution as a violin plot.
In the supplementary information, we display these error distributions 
of the general data set all combined, or separated per site or clock model.
The certainty in the model selection is again shown per parameter setting,
as a stacked bar. 
\richel{
  I am unsure how to depict that stacked bar and its uncertainty.
  Sure, the stacked bars (without uncertainty) will work,
  but information about uncertainty is lost. 
  See figure 'Showing a stacked bar with uncertainty'
  for three equivalent ways to display a stacked bar with uncertainty.
}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Supplement}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $b_g$                 & Speciation initiation rate of a good species & derived, see \ref{table:pbd_parameters} \\
    $b_i$                 & Speciation initiation rate of an incipient species & derived, see \ref{table:pbd_parameters} \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^9$ \\
    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2 \\
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_s$                 & Sampling method & S, L, R \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line is 
    the biological parameter set. 
    The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
    and so on.  
    The RNG seeds for the data set investigating the effect 
    of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
    The sampling methods are abbreviated as such: 'R' denotes random
    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
    data set. For the data set exploring the effect of sampling, we use 'shortest'
    and 'longest' for each value of $R_i$ (which are random seeds in which sampling has an effect).
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
